A Mountaintop View Requires Minimal Sorting: \break An Instance Optimal Contour Tree Algorithm

### 2014-10-30 paint-contour-tree.tex
typos corrected

Suppose $\MM$ has $n$ vertices, $N$ facets, and $t$ critical points.
A classic result of Carr, Snoeyink, and Axen (2000) give an algorithm that takes $O(n\log n + N\alpha(N))$ time
(where $\alpha(\cdot)$ is the inverse Ackermann function). A further improvement to $O(t\log t + N)$ time
was given by Chiang et al. All these algorithms involve a global sort of the critical points,
which is a major bottleneck for large datasets. Unfortunately, lower bounds of $\Omega(t\log t)$ also exist.

@@

A large class of data, such as satellite terrain maps, physical simulation output, and combustion data, can
be thought as a function $f: \RR^d \mapsto \RR$. Typically, a finite representation is given
by considering $f$ to be piecewise linear over some mesh (simplicial complex) $\MM$ in $\RR^d$.
\emph{Contour trees} are a specific topological data structure used to represent and visualize
the function $f$.

@@

\begin{figure}[h]\centering
    \includegraphics[width=.27\linewidth]{figs/spike3}%
    \hspace{.35in}
    \includegraphics[width=.27\linewidth]{figs/spike4}%
    \hspace{.4in}
    \includegraphics[width=.22\linewidth]{figs/spikeTree2}
    \caption{Two surfaces with different orderings of the maxima, but the same contour tree.}
    \label{fig:relative}
\end{figure}

@@

% \begin{figure}[h]\centering
%     \includegraphics[width=.3\linewidth,height=.42\linewidth]{figs/shaded}%
%     \hspace{.5in}
%     \includegraphics[width=.3\linewidth,height=.42\linewidth]{figs/trees2}
%     \caption{On the left, a surface will a balanced contour tree, but whose join and split trees have long tails.  
%     On the right (in clockwise order from the top) the contour, split and join trees.}
%     \label{fig:sorting}
% \end{figure}
\begin{figure}[h]\centering
    \includegraphics[width=.7\linewidth,height=.28\linewidth]{figs/balanced2}
    \includegraphics[width=.7\linewidth,height=.25\linewidth]{figs/trees}
    \caption{On top, a surface will a balanced contour tree, but whose join and split trees have long tails.  
    On bottom (from left to right), the contour, join and split trees.}
    \label{fig:sorting}
\end{figure}

@@

\Ben{This figure will not show up in black and white printout.}

>>>>

Suppose $\MM$ has $n$ vertices, $N$ facets, and $t$ critical points.
A classic result of Carr, Snoeyink, and Axen (2000) gives an algorithm that takes $O(n\log n + N\alpha(N))$ time
(where $\alpha(\cdot)$ is the inverse Ackermann function). A further improvement to $O(t\log t + N)$ time
was given by Chiang et al. All these algorithms involve a global sort of the critical points,
which is a major bottleneck for large datasets. Unfortunately, lower bounds of $\Omega(t\log t)$ also exist.

@@

A large class of data, such as satellite terrain maps, physical simulation output, and combustion data, can
be thought of as a function $f: \RR^d \mapsto \RR$. Typically, a finite representation is given
by considering $f$ to be piecewise linear over some mesh (simplicial complex) $\MM$ in $\RR^d$.
\emph{Contour trees} are a specific topological data structure used to represent and visualize
the function $f$.

@@

\begin{figure}[h]\centering
    \includegraphics[width=.27\linewidth]{figs/spikey1}%
    \hspace{.35in}
    \includegraphics[width=.27\linewidth]{figs/spikey2}%
    \hspace{.4in}
    \includegraphics[width=.22\linewidth]{figs/spikeyTree}
    \caption{Two surfaces with different orderings of the maxima, but the same contour tree.}
    \label{fig:relative}
\end{figure}

@@

% \begin{figure}[h]\centering
%     \includegraphics[width=.3\linewidth,height=.42\linewidth]{figs/shaded}%
%     \hspace{.5in}
%     \includegraphics[width=.3\linewidth,height=.42\linewidth]{figs/trees2}
%     \caption{On the left, a surface will a balanced contour tree, but whose join and split trees have long tails.  
%     On the right (in clockwise order from the top) the contour, split and join trees.}
%     \label{fig:sorting}
% \end{figure}
\begin{figure}[h]\centering
    \includegraphics[width=.7\linewidth,height=.3\linewidth]{figs/balanced3}
    \vspace{.1in}

    \includegraphics[width=.7\linewidth,height=.25\linewidth]{figs/trees}
    \caption{On top, a surface will a balanced contour tree, but whose join and split trees have long tails.  
    On bottom (from left to right), the contour, join and split trees.}
    \label{fig:sorting}
\end{figure}

@@
